Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-21 analysis/paper/paper.Rmd
added KS content

# set default theme for figures
theme_set(theme_bw())

@@

Four different mark-recapture abundance estimators were employed to explore variation in estimates. Estimators fell into two broad categories: single-census and multiple-census. For the single-census estimators, we used the Lincoln-Peterson estimator:

@@

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple-census estimator, we first utilized the Schnabel estimator:

@@

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) marking a fish does not affect its chances of recapture, (3) no loss of marks, (4) marks are not missed or mistaken, and (5) all fish have equal chance of being caught in the second (and subsequent) sample(s).

@@

```{r read-bio-outputs}
fmt_bio_out = function(x) {
  x = x %>%
    clean_names() %>%
    mutate(date = as.Date(date, "%m/%d")) %>%
    mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>% 
    mutate(month = month(date))
  return(x)
}

@@

To estimate the total biomass of fish potentially consumed by Northern Pikeminnow in Deadwater Slough, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the time period by the estimated population size for that time period. <!-- from which estimator? -->The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 10.9 g for the spring and 10.3 g for the fall, an average from juvenile Chinook Salmon captured during those date ranges at seven rotary screw traps located throughout the Upper Salmon upstream of Deadwater Slough.

@@

@@

***Finish spring abundance results.*** <!-- Can we assume equal capture probability and perhaps using the ratio of the average fall CPUE and spring CPUE. Then multiply the average fall abundance estimate by the ratio. Does that seem appropriate? -->

@@

@@

<!--If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.-->

@@

```{r, eval = F}
# pull in SAR numbers
sar = read_csv(here("analysis/data/raw_data", 
                    "WSSCh_sar_4See.csv")) %>%
  summarise(across(c(medsar, sdsar),
                   median)) %>%
  summarise(across(c(medsar, sdsar),
               ~ . / 100))

adult_impact = fall_only_no_growth %>%
  filter(cum_cons_fish_g == max(cum_cons_fish_g)) %>%
  select(cum_cons_fish_g) %>%
  mutate(n_chnk_smolts = cum_cons_fish_g / 12) %>%
  bind_cols(all_est %>%
              select(event_name, 
                     model,
                     N,
                     Lci,
                     Uci)) %>%
  mutate(across(c(smolt_N = N, 
                  smolt_Lci = Lci, 
                  smolt_Uci = Uci),
                ~ . * n_chnk_smolts)) %>%
  bind_cols(sar) %>%
  mutate(adult_N = smolt_N * medsar,
         adult_Lci = smolt_Lci * qnorm(0.025, medsar, sdsar),
         adult_Uci = smolt_Uci * qnorm(0.975, medsar, sdsar))

@@

@@

*	SAR results. Maybe a plot of multiple scenarios depending on gastric lavage results? And pop estimates if they vary in time.

@@

@@

@@

```{r cpue-fig, fig.cap = "The number of Northern Pikeminnow caught per angler effort for each sampling event."}
total_cpue %>%
  ggplot(aes(x = event_name,
             y = catch_per_hr)) +
  geom_col(fill = "black") +
  labs(y = "Catch Per Angler Hour") +
  theme(axis.title.x = element_blank()) +
  theme_classic()

# total_cpue %>%
#   ggplot(aes(x = event_name,
#              y = catch_lg_per_angler_day,
#              fill = event_name)) +
#   geom_col() +
#   scale_fill_brewer(palette = "Set1",
#                     name = "Event") +
#   labs(x = "Event",
#        y = "CPUE (Large Pikeminnow caught per angler day)") +
#   theme(legend.position = "none")

@@

```{r abund-fig, fig.cap = "Estimates of abundance of Northern Pikeminnow using different mark-recapture estimators and for the fall sampling events. Note the y-axis is truncated at 100."}
all_est %>%
  mutate(across(c(N, Lci, Uci),
                ~ . / 1000)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.1) +
  facet_wrap(~ event_name) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = "Estimator",
       y = "Abundance (x 1,000)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  scale_color_brewer(palette = "Dark2")

# all_est %>%
#   mutate(across(c(N, Lci, Uci),
#                  ~ . / 1000)) %>%
#   ggplot(aes(x = model,
#              y = N,
#              color = model)) +
#   geom_errorbar(aes(ymin = Lci,
#                     ymax = Uci),
#                 width = 0.1) +
#   geom_point(size = 4) +
#   facet_wrap(~ event_name) +
#   scale_color_brewer(palette = "Set1") +
#   coord_cartesian(ylim = c(0, 100)) +
#   labs(x = "Estimator",
#        y = "Abundance (x 1,000)") +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 45,
#                                    hjust = 1))

@@

```{r fall-bio-fig, eval = F, fig.cap = "First bioenergetics plot."}
fall_bio_p = fall_only_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Scumulative fish eaten (g)",
                     breaks = seq(0, 200, by = 10),
                     sec.axis = sec_axis(~. / 12, 
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 20, by = 1))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)")

fall_bio_p

```{r full-yr-ng-fig, eval = F, fig.cap = "Next bioenergetics plot."}
full_yr_ng_p = full_yr_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

full_yr_ng_p

```{r full-yr-10p-fig, eval = F, fig.cap = "Another bioenergetics plot."}
full_yr_10p_p = full_yr_10p_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 40, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

full_yr_10p_p

```{r alt-mod-fig, eval = F, fig.cap = "Another bioenergetics plot, again."}
alt_mod_p = alternative_mod %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

alt_mod_p

@@

```{r adult-imp-p, eval = F, fig.cap = "The adult impacts plot."}
adult_impact %>%
  mutate(stage = "Smolt") %>%
  select(event_name, model,
         stage,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(event_name, model,
                     stage,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  filter(str_detect(model, "Schuma", negate = T)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.5) +
  geom_point(size = 4) +
  facet_grid(stage ~ event_name,
             scales = "free_y") +
  scale_color_brewer(palette = "Set1",
                     name = "Model") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(x = "N. Pikeminnow Abundance Model")
```

>>>>

# set default theme for figures
#theme_set(theme_bw())
theme_set(theme_classic())

@@

Four different mark-recapture abundance estimators were employed to explore variation in estimates among them. Estimators fell into two broad categories: single-census and multiple-census. For the single census estimators, we treated the first week of sampling as the mark event, and the following (second) week as the recapture week, pooling data within each of those weeks. Alternatively, the multiple census estimators treat each day as a sampling occassion, and use information about the total marked fish from all previous occassions to infer the total abundance.

For the single-census estimators, we first used the Lincoln-Peterson estimator:

@@

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
For a multiple-census estimator, we first utilized the Schnabel method:

@@

All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the sampling period), (2) all fish have equal chance of being caught in the second (and subsequent) sample(s), (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, (5) marks are not missed or mistaken.

@@

```{r read-bio-outputs, message=FALSE, warning=FALSE}
fmt_bio_out = function(x) {
  x = x %>%
    clean_names() %>%
    mutate(date = as.Date(date, "%m/%d")) %>%
    mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>% 
    mutate(month = month(date))
  return(x)

}

@@

To estimate the total biomass of fish potentially consumed by Northern Pikeminnow in Deadwater Slough, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the timer period by the estimated population size for that time period (mean Schnabel estimates in the fall, CPUE ratio generated estimate in the spring). The total biomass was further converted to an estimate of the total number of Chinook salmon consumed by dividing by a weight of `r avg_DSR_g` for the fall and `r avg_NRR_g` for the spring, an average from juvenile Chinook Salmon captured during those date ranges at seven rotary screw traps located throughout the Upper Salmon upstream of Deadwater Slough.

@@

```{r adult-impacts}
# pull in SAR numbers
sar = read_csv(here("analysis/data/raw_data", 
                    "WSSCh_sar_4See.csv")) %>%
  summarise(across(c(medsar, sdsar),
                   median)) %>%
  summarise(across(c(medsar, sdsar),
               ~ . / 100))

# summarize potential Chinook eaten for each time period by one pikeminnow
adult_impact = fall_results %>%
  summarise(across(ends_with("g_n"), max)) %>%
  add_column(season = "Fall",
             .before = 0) %>%
  bind_rows(spring_results %>%
              summarise(across(ends_with("g_n"), max)) %>%
              add_column(season = "Spring",
                         .before = 0)) %>%
  pivot_longer(cols = ends_with("g_n"),
               names_to = "perc_diet",
               values_to = "juv_chnk") %>%
  mutate(across(perc_diet,
                str_extract,
                "[:digit:]+"),
         across(perc_diet,
                as.integer)) %>%
  left_join(spring_abund %>%
              select(N_F:Uci_F,
                     N_S:Uci_S) %>%
              pivot_longer(everything()) %>%
              mutate(season = if_else(str_detect(name, "_F$"),
                                      "Fall",
                                      "Spring")) %>%
              mutate(across(name,
                            str_remove,
                            "_F$"),
                     across(name,
                            str_remove,
                            "_S$")) %>%
              pivot_wider(names_from = name,
                          values_from = value)) %>%
  mutate(smolt_N = N * juv_chnk,
         smolt_Lci = Lci * juv_chnk,
         smolt_Uci = Uci * juv_chnk) %>%
  bind_cols(sar) %>%
  mutate(adult_N = smolt_N * medsar,
         adult_Lci = smolt_Lci * qnorm(0.025, medsar, sdsar),
         adult_Uci = smolt_Uci * qnorm(0.975, medsar, sdsar))

# sum fall and spring impacts
total_impacts = adult_impact %>%
  group_by(perc_diet) %>%
  summarise(across(c(ends_with("_N"),
                     ends_with("_Lci"),
                     ends_with("_Uci")),
                   sum))

@@

All the estimators yielded estimates with overlapping confidence intervals. Because the sampling design most closely matched a multiple census estimator, we feel the Schnabel estimator is the most appropriate. We examined diagnostic plots for that estimator (number of marks vs. proportion of recaptures) and they suggested that we were meeting the assumptions of the Schnabel estimator. Therefore, further results will be based on the Schnabel estimates of abundance.

Using the mean of the Schnabel estimates of Northern Pikeminnow abundance across two fall sampling events, and the mean of the CPUE from those events as well, we estimated there to be `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95%CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough during spring of 2020.

@@

The estimated number of Chinook Salmon consumed by Northern Pikeminnow begins to taper off between 30-90% of Pikeminnow diet. So although we are uncertain of that diet proportion, the impact of that uncertainty is somewhat constrained. We estimated the total juvenile outmigrants consumed (fall and spring) to be between `r paste(prettyNum(round(range(total_impacts$smolt_N)), big.mark = ","), collapse = " and ")`. This translates to adult Chinook equivalents between `r paste(prettyNum(round(range(total_impacts$adult_N)), big.mark = ","), collapse = " and ")` (Figure \@ref(fig:adult-imp-p)).

@@

All estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. <!-- This might be a good place to bring in some comparisons with Northern Pikeminnow densities from elsewhere to put our numbers in context.-->

If the first mark-recapture assumption was violated and the population is actually open (potential immigration and emigration during the survey period), as long as the immigration and emigration rates are equal between marked and unmarked fsih, our estimators should still provide unbiased estimates of abundance. Given the size of the sampling area, and the length of the sampling period (two weeks).

We are comfortable with assumptions 4 and 5 (no loss of marks and no mistaken or missed marks). Assumptions 2 and 3 indicate whether the second (and subsequent) surveys are random samples of the population. Angling methods may provide a size selection bias, but that may only affect the interpretation of the abundance estimates (i.e., our estimate are the abundance of Northern Pikeminnow above some size threshold). Unequal catchability between individual fish is very difficult to assess, especially if that heterogeneity arises from being caught once.

<!-- If catching and marking fish makes them want to leve the Deadwater area (more than unmarked fish), that could be making our estimates biased high (due to fewer recaptures). I don't imagine this is happening, but does anyone else have concerns about that? -->

The spring abundance estimate is smaller than either fall sampling event. However, our approach assumes equal capture probabilities betwee the fall and spring sampling events. If it does differ, we believe that the capture probability in the spring could be lower due to higher flows, which would lead to our spring abundance estimate being biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.

@@

@@

Finally, we estimated the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough reach was reduced or eliminated by multiplying the total estimated consumed Chinook Salmon by an estimate of Granite-to-Granite smolt-to-adult survival rates (SARs) (`r round(sar$medsar, 5)`, SE: `r round(sar$sdsar, 5)`) [@McCann2019].

@@

*	SAR results. Those results are GRA - GRA SARs, estimated from multiple years of PIT tag data. Seem pretty good to me.
* Maybe a plot of multiple scenarios depending on gastric lavage results? And pop estimates if they vary in time.
* Maybe pop estimates if they vary in time.

@@

```{r spring-N}
spring_abund %>%
  rowwise() %>%
  mutate(`Avg. Fall N (CI)` = paste0(prettyNum(round(N_F), big.mark = ","),
                                     " (",
                                     paste0(prettyNum(round(c(Lci_F, Uci_F)), big.mark = ","),
                                            collapse = ", "),
                                     ")"),
         `Spring N (CI)` = paste0(prettyNum(round(N_S), big.mark = ","),
                                     " (",
                                     paste0(prettyNum(round(c(Lci_S, Uci_S)), big.mark = ","),
                                            collapse = ", "),
                                     ")")) %>%
  ungroup() %>%
  select(Species = species,
         `Avg Fall CPUE` = cpue_F,
         `Spring CPUE` = cpue_S,
         ends_with("(CI)")) %>%
  kable(booktabs = T,
        caption = "Average CPUE and abundance in the fall (95% confidence interval), and CPUE and estimates of spring abundance (95% confidence interval) for Northern Pikeminnow in Deadwater Slough.") %>%
  kable_styling()

@@

```{r adult-imp}
adult_impact %>%
  mutate(stage = "Smolt") %>%
  select(#event_name, 
         #model,
         stage,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(#event_name, model,
                     stage,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        col.names = c("Stage",
                      "N",
                      "Lower 95% CI",
                      "Upper 95% CI"),
        digits = 0,
        caption = "Estimates of smolts consumed during fall migration by Northern Pikeminnow, and the adult equivalent of those smolts based on SARs.") %>%
  kable_styling()
```

@@

```{r cpue-fig, fig.cap = "The number of Northern Pikeminnow caught per angler effort for each sampling event."}
total_cpue %>%
  ggplot(aes(x = event_name,
             y = catch_per_hr)) +
  geom_col(fill = "gray30") +
  labs(y = "Catch Per Angler Hour") +
  theme(axis.title.x = element_blank())

@@

```{r abund-fig, fig.cap = "Estimates of abundance of Northern Pikeminnow using different mark-recapture estimators and for the fall sampling events. Note the y-axis is truncated at 100."}
all_est %>%
  mutate(across(c(N, Lci, Uci),
                ~ . / 1000)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.1) +
  facet_wrap(~ event_name) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = "Estimator",
       y = "Abundance (x 1,000)") +
  #theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  scale_color_brewer(palette = "Dark2")

@@

@@

```{r adult-imp-p, fig.cap = "How many Chinook Salmon are consumed by Northern Pikeminnow, depending on what percent of diet is Chinook. Faceted by migration season and life-stage."}
adult_impact %>%
  mutate(stage = "Juv. Emigrant") %>%
  select(stage,
         season,
         perc_diet,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(stage,
                     season,
                     perc_diet,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  ggplot(aes(x = perc_diet,
             y = N,
             color = as.factor(perc_diet))) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.1) +
  geom_point(size = 4) +
  facet_grid(fct_rev(stage) ~ season,
             scales = "free_y") +
  scale_color_viridis_d(name = "Percent Chinook\nin Diet") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "Number Chinook Consumed")
